Agda Meets Accelerate

### 2012-07-31 paper/main.tex
Revised and extended the Accelerate subsection

Accelerate \cite{ChakravartyKellerLeeMcdonellGrover2011} is an
embedded DSL for running data parallel code on a GPU.

Accelerate has a rich typed interface that already encodes many
invariants using advanced typing concepts in Haskell, in particular
GADTs\cite{PeytonJonesVytiniotisWeirichWashburn2006}, associated types
\cite{ChakravartyKellerJones2005}, and type functions
\cite{SchrijversPeytonJonesChakravartySulzmann2008}.

Moreover, Accelerate is a \emph{generative library}, in the sense that
most of its API operations just construct abstract syntax trees (AST). In the end,
the \verb+run+ operation accepts such an AST (of type \verb+Acc a+),
compiles it to GPU kernels, 
uploads it to the device, executes it, and retrieves the results. 
\begin{verbatim}
CUDA.run :: Arrays a => Acc a -> a
\end{verbatim}
The type class constraint \verb+Arrays a+ restricts the result type to
a type that can be handled by accelerate, usually an array type.

A simple Accelerate program would look like this:
\begin{verbatim}
dotp :: Vector Float -> Vector Float -> Acc (Scalar Float)
dotp xs ys = let xs' = use xs
                 ys' = use ys
             in  fold (+) 0 (zipWith (*) xs' ys')
\end{verbatim}
Here, \verb+Vector+ and \verb+Scalar+ are one- and zero-dimensional
array types known to Accelerate. Initially, the values of the
\verb+Vector+ type are simply Haskell arrays. The \verb+use+ operation
transforms a Haskell array into an array reference on the GPU and
makes the reference available for use in constructing the AST of the
desired operation. The last line applies two API operations,
\texttt{fold} and \texttt{zipWith} to construct an AST that returns a
\texttt{Scalar Float}, i.e., a single float value. This line relies
heavily on (type class) overloading: \texttt{0}, \texttt{(+)}, and
\texttt{(*)} are overloaded to just construct abstract syntax.

\textbf{TODO}: explain type \texttt{Array sh a}, in particular the
\texttt{sh} part.

>>>>

Accelerate~\cite{ChakravartyKellerLeeMcdonellGrover2011} is a data-parallel array language embedded into Haskell, which targets GPUs. It is a \emph{generative library}, as its data-parallel array operations are not executed directly, but instead construct abstract syntax trees (AST) representing an entire data-parallel subcomputation. These \emph{computation representations} are executed using a \verb+run+ operation that accepts such an AST (of type \verb+Acc a+), compiles it to GPU kernels, uploads it to the device, executes it, and retrieves the results.
%
\begin{verbatim}
CUDA.run :: Arrays a => Acc a -> a
\end{verbatim}
%
The type class constraint \verb+Arrays a+ restricts the result type to a single array or a tuple of arrays.

As computation representations of type \verb+Acc a+ are compiled at application runtime, all \verb+Acc+ compilation errors are effectivly \emph{runtime errors} of the application. Hence, Accelerate uses a range of Haskell type system extensions to statically type Accelerate expressions, such that these runtime errors are avoided where possible. In particular, Accelerate uses GADTs\cite{PeytonJonesVytiniotisWeirichWashburn2006}, associated types
\cite{ChakravartyKellerJones2005}, and type families
\cite{SchrijversPeytonJonesChakravartySulzmann2008}.

As a simple example of an Accelerate program, consider a function implementing a dot product:
%
\begin{verbatim}
dotp :: Vector Float -> Vector Float -> Acc (Scalar Float)
dotp xs ys = let xs' = use xs
                 ys' = use ys
             in  fold (+) 0 (zipWith (*) xs' ys')
\end{verbatim}
%
The types \verb+Vector+ and \verb+Scalar+ represent one- and zero-dimensional
arrays. Plain arrays, such as \verb+Vector Float+ are conventional Haskell arrays, using an unboxed representation to improve performance. However, when they are wrapped into the constructor \verb+Acc+, such as in \verb+Acc (Scalar Float)+, they represent arrays of the embedded language and are allocated in GPU memory, which in current high-performance GPUs is physically separate from CPU memory.

The \verb+use+ operation makes a Haskell array available in the embedded language by wrapping it into the \verb+Acc+ constructor and copying it to GPU memory.\footnote{Accelerate employs a caching strategy to avoid the transfer of arrays, which are already available in GPU memory.} The operations \texttt{fold} and \texttt{zipWith} represent collective operations on Accelerate arrays, effectively producing a representation of an array computation yielding a value of
\texttt{Scalar Float}; i.e., a single float value. This code relies
heavily on (type class) overloading: \texttt{0}, \texttt{(+)}, and
\texttt{(*)} are overloaded to just construct abstract syntax.

The type \verb+Scalar+ and \verb+Vector+ are merely type synonyms instantiating a shape-parameterised array type to the special case of zero and one dimensional arrays:
%
\begin{verbatim}
type Scalar e = Array DIM0 e
type Vector e = Array DIM1 e
\end{verbatim}
%
So, the general type for \verb+use+ is
%
\begin{verbatim}
use :: Elt e => Array sh e -> Acc (Array sh e)
\end{verbatim}
%
where the class \verb+Elt+ characterises all types that may be held in Accelerate arrays. These are currently primitive types and tuples.

Common dimensions, such as \verb+DIM0+, \verb+DIM1+, and so on, are predefined, but to enable shape polymorphic computations, along the lines pioneered in the Haskell array library Repa~\cite{keller-etal:repa}, shapes are inductively defined using type-level snoc lists formed from
%
\begin{verbatim}
data Z       = Z
data sh :. i = sh :. i
\end{verbatim}
%
Hence, we have got
%
\begin{verbatim}
type DIM0 = Z
type DIM1 = DIM0 :. Int
<and so on>
\end{verbatim}
